---
title: "Network Analysis in R (part1)"
author: "Sckinta"
date: "2019-07-11"
categories: ["r"]
tags: ["network"]
output: 
  blogdown::html_page:
          toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Network analysis is to study the complexity of the inter-relationships between actors of all sorts and provides an architectural view of individual actor connections. It has been applied to many fields, like social network and gene network, and useful for any systematic studies on individual relationship. [wiki](https://en.wikipedia.org/wiki/Graph_theory)

This post plus a few following posts aim to introduce the termilogy that are frequently used in network and public available R tools that are useful for network analysis and visualization.

# Glossaries

**Vertex**: is the node of network  
**Edge**: the connection of notes, sometimes called "linked". The edge can be **directed** or **undirected**, depending on whether the distinction between source and target is meaningful.  
**Graph**: constitute of vertex and edges, which represents the whole inter-relationship of nodes. The **subgraph**, from the other hand, represents the inter-relationship of a subset of nodes.  
**Cliques**: Fully connected subgraphs of a graph in which the every vertex connects with every other vertex.  
**Degree** : the number of adjacent vertex.  
**Order** : how far away from one vertex to another. order 0 is always v itself, order 1 is v plus its immediate neighbors, order 2 is order 1 plus the immediate neighbors of the vertices in order 1, etc.  
**Attribute**: attributes are associated with either vertex or edge. By default, the `name` is the mandatory atrribute for vertex and edge. If the edges have a magnitude attribute, the graph is considered **weighted**.  
**Bipartite networks**: two-mode networks, in which there are two types of vertex/nodes and  connections are only allowed between different types of nodes. Thus, `type` is a mandatory vertex attribute for bipartite networks. There are two types of common used bipartite networks, 1) the network consists of two types of vertex as individual and groups in which indiviudual (the first type of vertex) are members of group (the second type of vertex). This is also called "affilication network". 2) the network consistent of two "parallel" groups of vertex, in which group 1 only interact with group 2 while no interactions happen within the same group.

# IGRAPH object
**IGRAPH object** is a R object defined in `igraph` package. 

```{r igraph, echo=FALSE, warning=F, message=F}
library(igraph)
make_graph(letters[1:10], directed = T)
```

As mentioned in glassory part, graph is made with *vertices* and *edges*, and can be *directed* or *undirected*. The first line of IGRAPH `IGRAPH 7352909 DN-- 10 5 -- ` means a graph with id `7352909` is directed (`D`) and named (`N`) containing `10` vertices and `5` edges. The second line `+ attr: name (v/c)` indicates the attributes associated with vertex (`v`) and edge (`e`). In this example, only vertex attributes `name` were listed. It is character variable (`c`). The third and fourth line of IGRAPH literally print what edges look like. In our case, since graph is directed, the arrow shows the direction of edge.  

It is worth to mention that name of edges can be listed by vertices name in a `from|to` format. In most cases, names are the default attributes for vertex and can be used just like index to extract information directly from IGRAPH. We will discuss this in detail in "graph subset" session in this post.  

```{r warning=FALSE, message=FALSE}
library(igraph)
library(tidyverse)
library(ggraph)
library(gridExtra)
```

# IGRAPH object "I/O"

## Create a graph
There are many ways to create graph from scratch. 

* Easy graph:`graph_from_literal`, `make_graph`  
* Create from user data: `graph_from_edgelist`, `graph_from_adjacency_matrix`, `graph_from_data_frame`
* Random graphs: `sample_gnp`, `sample_gnm`, `sample_pa`, `sample_smallworld`, etc.

Here we are going to focus on "create from user data" and some "random graphs" options. 

### From edge list 

The example below generates a directed graph from a list of vertex pair (edge)

```{r edgelist}
edge_list = data.frame(from = c(1, 2, 2, 3, 4), to = c(2, 3, 4, 2, 1))
edge_list
m = as.matrix(edge_list)
g = graph_from_edgelist(m,directed = TRUE)
g
plot(g)
```

If the edge is given in a single vector, the default make_graph will use the order of vector to make one-mode graph. It means that the first edge points from the first element to the second, the second edge from the third to the fourth, etc.

```{r make_graph_from_edgeVector}
g = make_graph(letters[1:10], directed = T)
g
plot(g)
```

### From adjacent matrix

The example below generates a undirected graph from a binary matrix, in which  nodes are matrix colname and rowname.

```{r adjacentMatrix}
set.seed(123)
adj_matrix = matrix(sample(0:1, 100, replace=TRUE, prob=c(0.8,0.1)), nc=10)
colnames(adj_matrix)=letters[1:10]
rownames(adj_matrix)=letters[1:10]
adj_matrix
g = graph_from_adjacency_matrix(adj_matrix, mode = "undirected", weighted = T)
g
plot(g)
```

There are other modes available with details at https://igraph.org/r/doc/graph_from_adjacency_matrix.html

### From data.frame

The graph generated from data.frame can add all attributes at once. Here is the example from official website. The attributes for nodes (actors) are `age` and `gender`, and edge (relationship) attributes include same.dept, friendship and advice

```{r graphDataFrame}
actors <- data.frame(
  name=c("Alice", "Bob", "Cecil", "David","Esmeralda"),
  age=c(48,33,45,34,21),
  gender=c("F","M","F","M","F"))
relations <- data.frame(
  from=c("Bob", "Cecil", "Cecil", "David","David", "Esmeralda"),
  to=c("Alice", "Bob", "Alice", "Alice", "Bob", "Alice"),
  same.dept=c(FALSE,FALSE,TRUE,FALSE,FALSE,TRUE),
  friendship=c(4,5,5,2,1,1), 
  advice=c(4,5,5,4,2,3)
  )
g <- graph_from_data_frame(relations, directed=TRUE, vertices=actors)
g
plot(g)
```

### Create random graph

Random graph generation is useful when it comes to simulation.  

* `gnp` and `gnm` generate random graph by assumming the same constant probability for every possible edge. The required parameters for `gnp` and `gnm` are 

** `gnp` --- the number of nodes/vertex and the constant probability
** `gnm` --- the number of nodes/vertex and the number of edges

```{r gnp}
# sample 10 vertex, every possible edge is created with the same constant probability 0.3
set.seed(12)
gr <- sample_gnp(10, 0.3)
gr
plot(gr)

# sample 10 vertex, create a 15-edge graph
set.seed(123)
gr <- sample_gnm(10, 15)
gr
plot(gr)
```

* Random scale free network, which means a network whose degree of nodes distribution follows a power law. `sample_pa` generates scale-free graphs according to the Barabasi-Albert model. We start with a single vertex and no edges in the first time step. Then we add one vertex in each time step and the new vertex initiates some edges to old vertices. The probability that an old vertex is chosen is given by $p(i) ~ k_i * power + zero.appeal$

```{r BA_random}
set.seed(123)
gr <- sample_pa(100, power = 2)
plot(g, vertex.label= NA, edge.arrow.size=0.02,vertex.size = 0.5)

plot(density(degree_distribution(gr)))
```

## Extract vertex and edge and their attributes

To get vertex list and their attributes from graph object, we use `V(graph)$"attribute_name"` to convert graph object to vector.  

First we generate graph object `g` using the actor relationship code from From data.frame. Then we get actor node (name), age (attribute 1) and gender (attribute 2)

```{r, echo=F}
actors <- data.frame(
  name=c("Alice", "Bob", "Cecil", "David","Esmeralda"),
  age=c(48,33,45,34,21),
  gender=c("F","M","F","M","F"))
relations <- data.frame(
  from=c("Alice", "Bob", "Cecil", "David","David", "Esmeralda"),
  to=c("Bob", "David", "Alice", "Alice", "Bob", "Alice"),
  same.dept=c(FALSE,FALSE,TRUE,FALSE,FALSE,TRUE),
  friendship=c(4,5,5,2,1,1), 
  advice=c(4,5,5,4,2,3)
  )
g <- graph_from_data_frame(relations, directed=TRUE, vertices=actors)
```
```{r, echo=T, eval=T}
V(g)$name
V(g)$age
V(g)$gender
```

We can also get all the vertex attributes to a data.frame using `as_data_frame()`
```{r}
igraph::as_data_frame(g, what = "vertices")
```

Similarily, to get edge list and their attributes from graph object, we use `E(graph)$"attribute_name"` to convert graph object to vector. OR using `as_data_frame()` to convert all edges to a data.frame

```{r edge_attr}
# edge attributes
E(g)$same.dept
E(g)$friendship

# edge data.frame
igraph::as_data_frame(g, what = "edges")
```

## Add/Delete vertex and edges

Vertices and edges can be added to existing graph by `add_()` or `+`. Please be aware that vertices of added new edges must be from known vertices already in the graph.  

```{r add}
# add vertices 
g %>% add_vertices(2, name=c("Lisa","Zack")) # the first argument is number of vertex
g + vertices(c("Lisa","Zack"))

# add connected edges (even number of vertices). The vertices must be from known vertices already in the graph 
g %>% add_edges(c("Alice","Bob"))
g + edge(c("Alice","Bob"))

# add paths. The vertices must be from known vertices already in the graph 
g + path("Alice","Bob","Cecil")

# If a new vertex needs to be added to current graph, using add graph method instead 
g + make_graph(c("Alice","Bob","Bob","Melisa"))
```

Delete can be done in a similar way.

```{r delete}
### remove the vertices whose age is younger than 30
vertex_df = g %>% igraph::as_data_frame(what="vertices") %>% 
        dplyr::tbl_df() %>% 
        dplyr::mutate(index=row_number()) %>% 
        dplyr::filter(age < 30)
# remove vertices by index number 
g %>% delete_vertices(vertex_df$index)
# remove vertice by name 
g %>% delete_vertices(vertex_df$name)


### remove the edges with friendship <= 1 
edge_df = g %>% igraph::as_data_frame(what="edges") %>% dplyr::tbl_df() %>% mutate(index=row_number()) %>% mutate(name=paste(from,to,sep="|")) %>% filter(friendship <= 1)
# remove vertice by index
g %>% delete_edges(edge_df$index)
# remove vertice by name 
g %>% delete_edges(edge_df$name)

```

The attributes of vertices and edges can be added or deleted to existing graph by `set_vertex_attr()`/`set_edge_attr()` or `delete_vertex_attr()`/`delete_edge_attr()`.  

```{r add_attr}
# add a new attr "relationship" for people in the same dept
edge_df = g %>% igraph::as_data_frame(what="edges") %>% 
  mutate(relationship=ifelse(same.dept,"collegue",NA))

g %>% set_edge_attr("relationship", which(!is.na(edge_df$relationship)), edge_df$relationship[!is.na(edge_df$relationship)])

```

## Merge graph

Graphs union/intersection is to merge two or more graphs into one graph using the shared the vertices. 

```{r merge}
g2 = graph_from_literal("David"-+"Charlie"+-+"Lisa",
                        "Lisa"+-+"David"+-"Jim",
                        "Zack"+-"Esmeralda"-+"Bob",
                        "Zack"+-"Charlie",
                        "Lisa"+-"Lisa",
                        "Bob"-+"Alice"+-"Esmeralda"
                        )

#### union graph
g3 = igraph::union(g,g2)
#### graph intersection
g4 = igraph::intersection(g,g2)

par(mfrow=c(2,2)) 
plot(g, edge.arrow.size=.4)
plot(g2, edge.arrow.size=.4)
plot(g3, edge.arrow.size=.4)
plot(g4, edge.arrow.size=.4)
```

## Induce subgraph

The subgraph can be induced by either vertex or edge names/index. The edge names are in the form `from|to`. 

```{r subgraph}
# induce a subgraph using a list of vertices
igraph::induced_subgraph(g, v=c("Alice","Bob","Cecil"))

# induce a subgraph using edges
igraph::subgraph.edges(g, c("Alice|Bob","David|Bob","Cecil|Alice"), delete.vertices = TRUE)

# induce a subgraph using edges attribute (friendship score stronger than 3)
e1 = E(g)[E(g)$friendship > 3]
subgraph.edges(g, e1, delete.vertices = TRUE)
```

## Relationship between known nodes

Given known graph, we sometimes want to know relationship between two known nodes. eg. are those two nodes adjacent? are those two nodes linked by a path? what are the path to connect them? 

```{r two_nodes}
# check two nodes adjacent. return boolean
are_adjacent(g3,"Jim", "Zack")

# pathes to connect 
all_shortest_paths(g3, "Jim","Zack")$res[[1]] # return a igraph.vs

all_simple_paths(g3, "Jim", "Zack") # return a list of igraph.vs

all_simple_paths(g3, "Jim", "Zack", mode = "all")  # regardless of direction
```


In next post, I will discuss how to use igraph package to measure and cluster network and use the igraph object to its fullest potential.